Comparison of long and short amplicon metabarcoding for determination of fungal species diversity and spatial turnover

### 2019-08-12 writing/MSA2019_presentation.Rnw
quality and length plots

\title{Comparison of long and short amplicon metabarcoding for determination of fungal species diversity and spatial turnover in an ectomycorrhizal West African woodland}
\author{\mbox{\textbf{Brendan Furneaux}\inst{1}} \and \mbox{Mohammad Bahram\inst{2}} \and \mbox{Nourou Yorou\inst{3}} \and \mbox{Anna Rosling\inst{1}} \and \mbox{Martin Ryberg\inst{1}}}
\institute{\inst{1}Uppsala University \and \inst{2}Swedish University of Agricultural Sciences \and \inst{3}University of Parakou}
\date{13 August, 2019}
\logo{\includegraphics[width=2.5cm]{Uppsala_University_logo}}

@@

<<setup, include=FALSE>>=
knitr::opts_chunk$set(echo = FALSE, device = "tikz")
library(drake)
library(magrittr)
library(tidyverse)
library(ggplot2)
library(scales)
library(ggtree)
library(castor)
theme_set(theme_bw(base_size = 24) + theme(legend.direction = "vertical"))
loadd(bc_corr, bc_corr2, unif_corr, unif_corr2, qstats_join, datasets, longtree,
      taxon_LSUcons_rdp)
reads <- qstats_join %>%
  mutate_at("Region", factor,
            levels = c("ITS2", "ITS1", "short", "ITS", "LSU", "long")) %>%
  mutate_at("Dataset", factor,
            levels = c("short-ion", "short-pacbio", "long-pacbio")) %>%
  group_by(Dataset, Plate, Region)
reads_subset <- reads %>%
  do(if (nrow(.) > 10000) sample_n(., 10000) else .)
reverselog_trans <- function(base = exp(1)) {
    trans <- function(x) -log(x, base)
    inv <- function(x) base^(-x)
    trans_new(paste0("reverselog-", format(base)), trans, inv, 
              log_breaks(base = base), 
              domain = c(1e-100, Inf))
}
@

@@

\begin{frame}{Bioinformatics}
  \begin{itemize}
    \item Redo RSII circular consensus basecalling using new algorithm (Arrow).
    \item Demultiplex and trim primers.
    \item Cut out regions (ITS1, ITS2, full ITS, LSU) using ITSx.
    \item Quality filter.
  \end{itemize}
\end{frame}

@@

% <<errorrate, fig.width = 12>>=
% ggplot(reads_subset, aes(x = erate,
%                   linetype = Dataset,
%                   group = paste(Dataset, Region),
%                   color = Region)) +
%   geom_line(aes(y = 1 - ..y..), stat = "ecdf") +
%   scale_x_continuous(name = "Expected number of errors per base",
%                      trans = reverselog_trans(10),
%                      limits = c(NA, 1e-5),
%                      oob = squish) +
%   scale_y_continuous(name = "Fraction passing") +
%   scale_color_viridis_d(option = "C") +
%   scale_linetype_discrete()
% @
\end{frame}

@@

% <<totalerror, fig.width = 12>>=
% ggplot(reads_subset, aes(x = eexp,
%                   linetype = Dataset,
%                   group = paste(Dataset, Region),
%                   color = Region)) +
%   geom_line(aes(y = 1 - ..y..), stat = "ecdf") +
%   scale_x_continuous(name = "Expected number of errors",
%                      trans = reverselog_trans(10),
%                      limits = c(NA, 0.1),
%                      oob = squish) +
%   scale_y_continuous(name = "Fraction passing") +
%   scale_color_viridis_d(option = "C") +
%   scale_linetype_discrete()
% @
\end{frame}

@@

\begin{frame}{Correlogram}
<<correlogram, fig.width = 12>>=
bind_rows(as.data.frame(bc_corr$mantel.res) %>%
            mutate(tdist = "same", metric = "bray"),
          as.data.frame(bc_corr2$mantel.res) %>%
            mutate(tdist = "diff", metric = "bray"),
          as.data.frame(unif_corr$mantel.res) %>%
            mutate(tdist = "same", metric = "unifrac"),
          as.data.frame(unif_corr2$mantel.res) %>%
            mutate(tdist = "diff", metric = "unifrac")) %>%
  mutate(Significant = `Pr(corrected)` < 0.05) %>%
  ggplot(aes(class.index, Mantel.cor, group = tdist, color = tdist, shape = Significant)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  # geom_line(aes(y = exp(predict(bc_fit)))) +
  scale_y_continuous(limits = c(-0.05, NA), name = "Spatial Distance correlation") +
  scale_color_discrete(name = "Year") +
  xlab("distance (m)") +
  facet_wrap("metric") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1))
@
\end{frame}

>>>>

\title{Comparison of long and short amplicon metabarcoding for determination of fungal species diversity and spatial turnover}
\subtitle{in an ectomycorrhizal West African woodland}
\author{\mbox{\textbf{Brendan Furneaux}\inst{1}} \and \mbox{Mohammad Bahram\inst{2}} \and \mbox{Nourou Yorou\inst{3}} \and \mbox{Anna Rosling\inst{1}} \and \mbox{Martin Ryberg\inst{1}}}
\institute{\inst{1}Uppsala University \and \inst{2}Swedish University of Agricultural Sciences \and \inst{3}University of Parakou}
\date{13 August, 2019}
\logo{\includegraphics[width=2.5cm]{Uppsala_University_logo}}

@@

\logo{}

<<setup, include=FALSE>>=
knitr::opts_chunk$set(echo = FALSE, device = "tikz", cache = TRUE)
library(drake)
library(magrittr)
library(tidyverse)
library(ggplot2)
library(scales)
library(ggtree)
library(castor)
theme_set(theme_bw(base_size = 24) + theme(legend.direction = "vertical"))
loadd(bc_corr, bc_corr2, unif_corr, unif_corr2, qstats_join, datasets, longtree,
      taxon_LSUcons_rdp)
reads <- qstats_join %>%
  mutate_at("Region", factor,
            levels = c("ITS2", "ITS1", "short", "ITS", "LSU", "long")) %>%
  mutate_at("Dataset", factor,
            levels = c("short-ion", "short-pacbio", "long-pacbio")) %>%
  group_by(Dataset, Plate, Region)
reads_subset <- reads %>%
  do(if (nrow(.) > 10000) sample_n(., 10000) else .)
reverselog_trans <- function(base = exp(1)) {
    trans <- function(x) -log(x, base)
    inv <- function(x) base^(-x)
    trans_new(paste0("reverselog-", format(base)), trans, inv, 
              log_breaks(base = base), 
              domain = c(1e-100, Inf))
}
@

@@

\begin{frame}{Bioinformatics}
  \begin{itemize}
    \item Redo RSII circular consensus basecalling using Sequel algorithm (Arrow).
    \item Demultiplex and trim primers.
    \item Cut out regions (ITS1, ITS2, full ITS, LSU) using ITSx.
    \item Quality filter - up to 3 expected errors.
  \end{itemize}
\end{frame}

@@

<<errorrate, fig.width = 12, cache.check = reads_subset>>=
filter(reads_subset, Tech == "PacBio") %>%
  ggplot(aes(x = eexp,
             group = paste(Dataset, Region),
             color = Dataset)) +
  geom_line(aes(y = 1 - ..y..), stat = "ecdf", size = 1) +
  scale_x_continuous(name = "Expected number of errors",
                     trans = reverselog_trans(10),
                     limits = c(NA, 1e-5),
                     oob = squish) +
  facet_wrap(~Region, ncol = 2, strip.position = "right") +
  scale_y_continuous(name = "Fraction passing") +
  # scale_color_viridis_d(option = "C") +
  geom_vline(aes(xintercept = 3), linetype = 2)
@
\end{frame}

@@

<<length, fig.width = 12, cache.check = reads_subset>>=
filter(reads_subset, Tech == "PacBio", eexp <= 3) %>%
  ggplot(aes(x = length,
             group = paste(Dataset, Region),
             color = Dataset, fill = Dataset)) +
  geom_area(stat = "density", size = 1, alpha = 0.5) +
  scale_x_log10(name = "Length (bp)",
                limits = c(50, NA),
                oob = squish) +
  facet_wrap(~Region, ncol = 2, strip.position = "right", scales = "free_y") +
  scale_y_continuous(name = "Fraction passing") +
  # scale_color_viridis_d(option = "C") +
  geom_vline(aes(xintercept = 3), linetype = 2)
@
\end{frame}

@@

\begin{frame}{Correlogram}
<<correlogram, fig.width = 12>>=
bind_rows(as.data.frame(bc_corr$mantel.res) %>%
            mutate(tdist = "same", metric = "bray"),
          as.data.frame(bc_corr2$mantel.res) %>%
            mutate(tdist = "diff", metric = "bray"),
          as.data.frame(unif_corr$mantel.res) %>%
            mutate(tdist = "same", metric = "unifrac"),
          as.data.frame(unif_corr2$mantel.res) %>%
            mutate(tdist = "diff", metric = "unifrac")) %>%
  mutate(Significant = `Pr(corrected)` < 0.05) %>%
  ggplot(aes(class.index, Mantel.cor, group = tdist, color = tdist, shape = Significant)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.3, size = 1.5) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  # geom_line(aes(y = exp(predict(bc_fit)))) +
  scale_y_continuous(limits = c(-0.05, NA), name = "Spatial Distance correlation") +
  scale_color_discrete(name = "Year") +
  xlab("distance (m)") +
  facet_wrap("metric") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1))
@
\end{frame}
